The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2018-10-08 article.tex
Changes @Giappo

{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\gio{According to my fine graining approach we should at each step deepen every small section. At a certain level I think we can start to re-coarse-grain what we wrote to create the abstract.}
\richel{I enjoy this approach! Did some minor fine-graining}
\richel{Have you already looked up for a target journal? I know how a journal's constraints have helped me 
in writing an article, for example, by having a maximum number of pictures}

@@

\item There are many contemporary tools that provide the possibility to infer a phylogeny from genetic data (DNA, RNA, proteins). A popular Bayesian
phylogenetic tool is called BEAST and its cousin BEAST2.

\item BEAST is very flexible in setting up all possible phylogenetic priors (e.g. site/clock/speciation model).

\item Current limits in current tools.

@@

\item One of such speciation processes is the multiple birth hypothesis,
a new model (described below) and thus absent in BEAST.

\item The Multiple birth hypothesis can be useful to explain a phenomenon that has always puzzled evolutionary biologists: what are the drivers of the diversification processes for those phylogenies that show an impressive amount of speciation events in relatively short times? 
The (constant-rate) birth-death (BD) model embodies the common assumption that 
only a single speciation event can occur at any given time.
The multiple-birth-death (MBD) model \richel{I feel MBSD (Multiple Birth Single Death) may be a better name: extinctions are still one at a time} relaxes this assumption, allowing events in which 
large-scale environmental changes lead to a great number of species 
in relatively short time intervals. Such a hypothesis may be a better fit to describe the burst in cichlid fish diversification in systems like in the 
African Great Lakes: Malawi, Tanganyika and Victoria (\cite{janzen2016}, \cite{janzen2017}).

@@

\item Current phylogenetic tools assume that only a single speciation event can occur at any given time.
While this assumption is useful to construct a wide variety of successful 
models (for example: \cite{Maddison2007biSSE}, \cite{Valente2015}, \cite{etienne2012diversity}, \cite{etienne2014estimating}),
they disallow for environmental changes that trigger speciations in multiple clades at a same point in time.

@@

\item In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates. 
Additionally, $\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, all species present in the phylogeny at that moment
have a probability $q$ to speciate at that time, which is 
independent on $\lambda$.

@@

\item We start simulating 1000 \richel{I will measure the number of trees we'll be able to simulate within a short enough time, when the experiment is set up} MBD trees. From each MBD tree, a a DNA sequence alignment is simulated, after
which that alignment starts a Bayesian analysis. We use 
the 'pirouette' package (\cite{pirouette}) to call the BEAST2 tool 
suite from R. We let the Bayesian analysis assume a BD prior, to investigate
the error this inference makes due to this.

\item For each tree 
generated under the MBD model we aim to generate a "twin" tree under the BD model in order to perform a fair comparison, using trees 
with the same amount of information, 
which is the number of taxa 
and the same (expected) number of DNA mutations.
To obtain these twin trees, 
\richel{
  I suggest to first start with the equal number of taxa,
  and the calculation of the speciation rates first.
}
we impose
that the expected number of mutations in an MBD tree, $m_{MBD}$ equals
the expected number of mutations in a BD tree, $m_{BD}$:

@@

The expected number of mutations $m$ of a phylogeny 
\richel{
  I think 'expected number of mutations' would be more correct.
  Do you agree?
}
with crown age $T$ in fact is given by
\richel{
  above stood 'crown age -T'. 
  I feel that a crown age is a positive number,
  but I know you have had a good reason.
  Perhaps better would be to write something explicit like:
  t\_now - t\_crown = T.
  Looking forward for a better suggestion than mine :-)
}

@@

%It can be easily seen considering that $\rho$ is the number of mutations
%per unit time per site. For this reason it's needed to multiply by
%time and number of sites. 
\gio{This feels kind of a repetitions of what we wrote before the formula. I comment it. We can think of reinsert it afterwards, if needed (see comment above).}
\richel{
  I suggest to remove such commented-out lines. 
  Although I sometimes get attached to my sentences,
  they clog up the document by non-info and 
  I usually delete them anyways in the end.
  I cannot remember ever regretting this (would I, 
  I could find it in the git history).
  Would you agree?
}

Since we cannot know $n_{BD}(t)$ before running simulations
we need to replace it with a proxy. 
For this reason we will use the average number of
species in time according to the BD model. 
It's well known that this is equal to \gio{insert proper citation}
\richel{
  I see you use angle bracket as a notation for the expected
  value. I usually see 'E(x)' as the expected value for 'x',
  and this is used at the beloved https://en.wikipedia.org/wiki/Expected\_value.
  What are the reasons you prefer the notation with the angle brackets?  
}

@@

%$$
%<n_{BD}(t)> = n(t=0) * exp[(mu - lambda) * t] (6)
%$$

%The relation to use to get the equivalent lambda should therefore be

%$$ 
%m_{MBD} 
%  = L * mu * integral[exp[(mu - lambda) * t] dt] 
%  = (L * mu * exp[(mu - lambda) * t]) / (mu - lambda) (7)
%$$

%Here everything is known but lambda. So solve the lambda and use that value.

\gio{My doubt is if we need to use $m_{MBD}$ for the single tree or the same
quantity averaged on the full MBD dataset $<m_{MBD}>$.
Do you think is better to use the
individual $m_{MBD}$ for each tree or the average across the whole
dataset?}
\richel{
  I think a per-tree calculation of the mutation rates 
  is the best we can do. As there was some noise between us above,
  I think the iteration after the next will allow us to get a better
  idea about this
}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{mbd.jpg}
  \caption{
    How to create twin trees and alignments. From a focal MBD tree, a twin tree is produced as 
    such: (1) estimate the $\lambda_{BD}$ to get the same expected number of tips, (2) simulate a BD tree with that amount of tips (discard trees with different number of tips), (3) estimate a mutation rate to get an alignment with the same expected number of mutations, (4) simulate alignments with that amount of mutations (discard those that don't, the picture shows an alignment that should be discarded) 
  }
\end{figure}

\item We explained how we set the parameters for each twin BD tree. Using this rules we generate a BD dateset. We repeat the analysis, producing alignments for each tree and subsequently using BEAST to produce a posterior for each of them.

\item Now we have two datasets of posteriors to compare, one for the BD model and one for the MBD model.

\item To compare the results for the two models we measure the inference error using the nLTT statistic between known/true tree and posterior/inferred trees.
\richel{
  I would love to describe this more concrete. For example, when do
  we say something has an effect? If we avoid making such judgements,
  how will we visualize?
}

\richel{I removed the Bayes factor text. It is useful when letting BEAST2
pick more/overly complex models and see if that more complex model fits the
data better (penalized by its increased complexity, similar to the AIC). It has
its uses, but I am unsure if we already want to discuss this now
or first focus on the proper tree twinning}

@@

\item
\richel{
  I guess you know I am a fan of the Open Science Framework,
  in which you first register you work before you do the experiment
  (note: I will do some small pilots to estimate the complete time
  of the experiment). I think it is the proper and superior science,
  which helps us against writing down bullshit stories after having
  obtained the results (e.g. 'We expected A and indeed found it!').
  It also helps me structure my work: first think
  deeply about the experiment, then do it (instead of the mixing
  up the two phases). What are your thoughts on that?
}

\item

>>>>

{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\gio{According to my fine graining approach we should at each step deepen every small section. At a certain level I think we can start to re-coarse-grain what we wrote to create the abstract.}
\richel{I enjoy this approach! Did some minor fine-graining}
\richel{Have you already looked up for a target journal? I know how a journal's constraints have helped me 
\gio{Honestly I have literally no idea how to select a good journal for this kind of article.}
in writing an article, for example, by having a maximum number of pictures}

@@

\item There are many contemporary tools that provide the possibility to infer a phylogeny from genetic data (DNA, RNA, proteins). A popular Bayesian
phylogenetic tool is called BEAST (\cite{beast}) and its cousin BEAST2 (\cite{beast2}).

\item BEAST is very flexible, providing the user with the option to set up all possible phylogenetic priors (e.g. site/clock/speciation model).

%Current limits in current tools.
\item However, currently available priors can be not suitable to analyze some specific datasets. 
%With this work we aim to test whether or not the implementation of a new prior model is beneficial to study a specific kind of diversification process.

@@

\item One of such speciation processes is the multiple birth hypothesis,
a new model (described below) and thus currently absent in BEAST.

\item The Multiple birth hypothesis can be useful to explain a phenomenon that has always puzzled evolutionary biologists: what are the drivers of the diversification processes for those phylogenies that show an impressive amount of speciation events in relatively short times? 
The (constant-rate) birth-death (BD) model embodies the common assumption that 
only a single speciation event can occur at any given time.
The multiple-birth-death (MBD) model 
\richel{I feel MBSD (Multiple Birth Single Death) may be a better name: extinctions are still one at a time}
\gio{I always had the same doubt. The problem is that a good name should be, at the same time, both descriptive and short. I'll think about that.}
relaxes this assumption, allowing events in which 
large-scale environmental changes lead to a great number of species 
in relatively short time intervals. Such a hypothesis may be a better fit to describe the burst in in systems like cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria (\cite{janzen2016}, \cite{janzen2017}).

@@

\item Current phylogenetic tools assume that only a single speciation event can occur at any given time.
While this assumption is useful to construct a wide variety of successful 
models (e.g \cite{Maddison2007biSSE}, \cite{Valente2015}, \cite{etienne2012diversity}, \cite{etienne2014estimating}),
they disallow for environmental changes that trigger speciations in multiple clades at a same point in time.

@@

\item In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates present also in the standard BD model. 
Additionally, MBD relies on two additional parameters. Parameter $\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, all species present in the phylogeny at that moment
have a probability $q$ to speciate at that time, which is 
independent on $\lambda$.

@@

\item We start simulating $N_{S} = 1000$ \richel{I will measure the number of trees we'll be able to simulate within a short enough time, when the experiment is set up} MBD trees. From each MBD tree, a DNA sequence alignment is simulated. For each sequence alignment we then perform a Bayesian analysis to recover a posterior distribution of trees, each composed of $N_{P}$ phylogenies. Such analysis is performed using 
the 'pirouette' package (\cite{pirouette}) to call the BEAST2 tool 
suite from R. We let the Bayesian analysis assume a BD prior, to investigate
the error this inference makes due to this.

\item For each tree 
generated under the MBD model we aim to generate a "twin" tree under the BD model in order to perform a fair comparison. 
We want trees from the two models to contain the same amount of information, 
i.e. the same (expected) number of DNA mutations. %option 1
%i.e. the same number of taxa at the present. %option 2
To obtain these twin trees, 
\richel{
  I suggest to first start with the equal number of taxa,
  and the calculation of the speciation rates first.
}
\gio{@richel: both assumptions seem to make sense to me. If possible I would run both and see which one performs better. For now let's stick to option 1 in the manuscript.}
we impose
that the expected number of mutations in an MBD tree, $m_{MBD}$ equals
the expected number of mutations in a BD tree, $m_{BD}$:

@@

The expected number of mutations $m$ of a phylogeny 
\richel{
  I think 'expected number of mutations' would be more correct.
  Do you agree?
}
\gio{I don't agree. It is expected if you use an expected number of species over time. If you have the correct $n(t)$ in principle you should be able to get the exact number of mutations.}
with crown age $-T$ (with $T>0$) in fact is given by
\richel{
  above stood 'crown age -T'. 
  I feel that a crown age is a positive number,
  but I know you have had a good reason.
  Perhaps better would be to write something explicit like:
  t\_now - t\_crown = T.
  Looking forward for a better suggestion than mine :-)
}
\gio{We can specify that the crown age is $-T$ where $T$ is a positive number (which should be already clear, I guess, if we report it as $-T$). In my opinion this makes everything nicer when we have to write down the integral which should, in principle, be $\int_{-T}^{0} f(t-(-T))$ but using integral properties can be rewritten as $\int_{0}^{T} f(t)$. Having a definite integral starting from $0$ always looks better ^^.}

@@

%It can be easily seen considering that $\rho$ is the number of mutations
%per unit time per site. For this reason it's needed to multiply by
%time and number of sites. 
\gio{This feels kind of a repetitions of what we wrote before the formula. I comment it. We can think of reinsert it afterwards, if needed (see comment above).}
\richel{
  I suggest to remove such commented-out lines. 
  Although I sometimes get attached to my sentences,
  they clog up the document by non-info and 
  I usually delete them anyways in the end.
  I cannot remember ever regretting this (would I, 
  I could find it in the git history).
  Would you agree?
}
\gio{Agree. Remove them. CLOSED}

Since we cannot know $n_{BD}(t)$ before running simulations
we need to replace it with a proxy. 
For this reason we will use the average number of
species in time according to the BD model. 
It's well known that this is equal to \gio{insert proper citation}
\richel{
  I see you use angle bracket as a notation for the expected
  value. I usually see 'E(x)' as the expected value for 'x',
  and this is used at the beloved https://en.wikipedia.org/wiki/Expected\_value.
  What are the reasons you prefer the notation with the angle brackets?  
}
\gio{Because the capital E is inelegant, plebeian and boor. And we want to be classy, don't we? :D}

@@

\gio{My doubt is if we need to use $m_{MBD}$ for the single tree or the same
quantity averaged on the full MBD dataset $<m_{MBD}>$.
Do you think is better to use the
individual $m_{MBD}$ for each tree or the average across the whole
dataset?}
\richel{
  I think a per-tree calculation of the mutation rates 
  is the best we can do. As there was some noise between us above,
  I think the iteration after the next will allow us to get a better
  idea about this
}
\gio{I am still very debated about this. I actually would like to ask also Rampal's opinion.
}
\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{mbd.jpg}
  \caption{
    How to create twin trees and alignments. From a focal MBD tree, a twin tree is produced as 
    such: (1) estimate the $\lambda_{BD}$ to get the same expected number of tips, (2) simulate a BD tree with that amount of tips (discard trees with different number of tips), (3) estimate a mutation rate to get an alignment with the same expected number of mutations, (4) simulate alignments with that amount of mutations (discard those that don't, the picture shows an alignment that should be discarded) 
  }
\end{figure}

\item We explained how we set the parameters for each twin BD tree. Using this rules we generate a BD dataset. We repeat the analysis, producing alignments for each tree and subsequently using BEAST to produce a posterior for each of them.

\subsection{Model selection}

\item So far we have simulated two datasets of trees under the two models: $\{T_{i}^{BD}\}_{i=1}^{N_{S}}$ and $\{T_{i}^{MBD}\}_{i=1}^{N_{S}}$.
We used them to generate a dataset of alignments for each model: $\{X^{BD}_{i}\}_{i=1}^{N_{S}}$ and $\{X^{MBD}_{i}\}_{i=1}^{N_{S}}$. From each dataset we produced a posterior distribution from a BD prior: 
$P_{i}(\theta | X^{BD}_{i}, BD)$ and $P_{i}(\theta | X^{MBD}_{i}, BD)$.
\gio{1) We might want to rename the models, e.g. BD = (0) and MBD = (1). These names with capital letters are too big and ugly;
2) Writing down things to tidy up manuscript and thoughts I just realized that actually the "model" is the same for the two posteriors. In fact with the word "model" Maturana et al. refer to the prior used (see eq. 1, $\pi(\theta | M)$). In such case are we sure we can actually use the Bayes factor for model selection?}
%Now we have two datasets  of posteriors to compare, one for the BD model and one for the MBD model.

\subsubsection{Option 1: nLTT statistics}
\item To compare the results for the two models we measure the inference error using the nLTT statistic between known/true tree and posterior/inferred trees (\cite{nltt}). To obtain such statistics the procedure is the following:

- From each tree $T_{i,j}^{M}$ (with $j=1,...,N_{S}$) belonging to the posterior $P_{i}(\theta | X^{M}_{i}, BD)$ and relative to the model $M$, we extrapolate the lineage-through-time (LTT), in other words we measure the number of species as a function of time $n_{i,j}(t)$. To allow a comparison we normalize dividing by the maximum number of species of each tree, i.e. the number of tips at the present $N_{i,j}(t)=\frac{n_{i,j}(t)}{n^{max}_{i,j}}$. We then define the nLTT measure as

$nLTT_{i,j} = \int_{0}^{T} \abs{N_{i,j}(t) - N_{T_{i}}} dt$

\gio{I am running out of letters :(}
\richel{
  I would love to describe this more concrete. For example, when do
  we say something has an effect? If we avoid making such judgements,
  how will we visualize?
}

\subsubsection{Option 2: Bayes Factor (BF)}

\richel{I removed the Bayes factor text. It is useful when letting BEAST2
pick more/overly complex models and see if that more complex model fits the
data better (penalized by its increased complexity, similar to the AIC). It has
its uses, but I am unsure if we already want to discuss this now
or first focus on the proper tree twinning}
\gio{But after we describe the tree twinning we will have to describe the method we use for model selection anyways. BF and nLTT are mutually exclusive choices, right?}
\gio{The point is: what method do we choose? According to the choice we have to describe that. I'd also like to know Rampal's opinion on that. In principle I like more the BF method cause it looks more solid while the nLTT looks a bit like the last resort to me.}

@@

\item
\richel{
  I guess you know I am a fan of the Open Science Framework,
  in which you first register you work before you do the experiment
  (note: I will do some small pilots to estimate the complete time
  of the experiment). I think it is the proper and superior science,
  which helps us against writing down bullshit stories after having
  obtained the results (e.g. 'We expected A and indeed found it!').
  It also helps me structure my work: first think
  deeply about the experiment, then do it (instead of the mixing
  up the two phases). What are your thoughts on that?
}
\gio{I agree on the principle but I don't fully agree on the open science procedure. Sometimes you do find something good while you're looking for something else. That happened many times in the history of science. This doesn't necessarily implies that we will end up p-hacking the results. I think we need to be open to new findings that might be not forecasted by our hypothesis. I think this is the good way to do science: be critical with our results but not imposing that reality will follow the theory.}
\item
